---
title: "Homework 11: Batch Processing"
author: "Chika Ikpechukwu."
date: "May 08, 2024"
output: 
  cleanrmd::html_document_clean:
    theme: tacit
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

### Preliminaries

**Open relevant libraries**

```{r eval = TRUE, echo = TRUE, message=FALSE, warning=FALSE}
library(kableExtra) # To make fancy tables
```

**Create a new folder for the original datasets.** 
(Note the code is commented out as 'add-folder()' requires unique filenames to run and the folder has already been created.)

```{r eval = TRUE, echo = TRUE, message=FALSE, warning=FALSE}

# source function files 
source("StrategicProgramming/barracudar/AddFolder.R") # create folders for an r project
# add_folder("OriginalData")
```

### Pseudocode:

1.    Get the file name
2.    Clean data for any empty/missing cases.
3.    Extract the year from each file name.
4.    Calculate abundance for each year (total number of individual birds found).
5.    Calculate species richness for each year (number of unique species found).

### User-Designed Functions
To execute above 5 functions were designed. The functions were:

1.    *getFilename()*: gets the filename for every file within a directory that contains 'countdata' in the filename 
2.    *clean_data()*: gets rid of any missing/empty cases in the dataframe using 'scientificName' as the unique ID.
3.    *extract_year()*: takes in a vector of filenames and returns the years embedded in each filename. 
4.    *abundance()*: takes in a 'cleaned' dataframe and returns the total number of individual birds observed.
5.    *species_richness()*: takes in a 'cleaned' dataframe and returns the number of unique species observed. 

See below chunk for the functions code: 
```{r eval = TRUE, echo = TRUE}


#############################################################
# FUNCTION: getFilename
# Description:  Get the filename for every file within a directory that contains 'countdata' in the filename
# Input(s):  'filePath' the directory path to be probed
# Output(s):  'filename' the filename with 'countdata'
# ------------------------------------------------------

getFilename <- function(filePath = NULL){

filename <- sub(".*/(.+\\.csv)$", "\\1", filePath)

return(filename)

} # end of getFilename
#############################################################
 
#############################################################
# FUNCTION: clean_data
# Description: data cleaning to get rid of any missing cases
# Input(s):  'filePath' the file path to the csv file to be cleaned
# Output(s): 'df' a clean (ie without missing values for the unique identifier) dataframe
# ------------------------------------------------------
clean_data <- function(filePath = NULL){
  d <- read.csv(filePath)
  df <- d[complete.cases(d$scientificName), ] # returns all columns for all rows with present 'scientificName'
  
  return(df)
  
} # end of clean_data

#############################################################

#############################################################
# FUNCTION: extract_year
# Description:  extract the years from a vector containing countdata filenames for various years
# Input(s):  'filePath' the filepath
# Output(s):  'year' the year of the file
# ------------------------------------------------------
extract_year <- function(filePath = NULL){
  
  year <- gsub(pattern = ".*\\.(\\d{4})-\\d{2}\\..*",
                replacement = "\\1",
                x = filePath)
    return(year)
  
} # end of extract_year
#############################################################

#############################################################
# FUNCTION: abundance
# Description:  count the number of individual birds observed
# Input(s):  'df' a dataframe
# Output(s):  'count' the number of individual birds observed
# ------------------------------------------------------
abundance <- function(df = NULL){
  
  count <- length(df$scientificName)
  
  return(count)
  
} # end of abundance
#############################################################

#############################################################
# FUNCTION: species_richness
# Description: count the number of unique birds observed
# Input(s):  'df' a dataframe
# Output(s):  'richness' count the number of unique birds observed 
# ------------------------------------------------------
species_richness <- function(df = NULL){
  
  richness <- length(unique(df$scientificName))
  
  return(richness)
  
} # end of species_richness
#############################################################

```

### Relevant BART Files for Analysis 
The filenames and full paths for every BART data file containing 'countdata' in the dile name was gotten using a for loop as shown below: 

```{r eval = TRUE, echo = TRUE}

# Using for loops, iterate through each year’s folders to gather the file names of these “countdata” .csv files.

rootDir <- "C:/Users/Chika Ikpechukwu/OneDrive - University of Vermont/Desktop/OriginalData"
dir_names <- list.dirs(path = rootDir, full.names = TRUE, recursive = FALSE) # returns all the folder names in the specified path.

Filenames <- c() #initiate empty vector for the filenames containing 'countdata'
Paths <- c() # initiate empty vector to store path to the 'countdata' files

for (i in seq_along(dir_names)){
  path <- list.files(path = dir_names[i],
                  pattern = "countdata",
                  full.names = TRUE) # get full file path of any file with "countdata" in a the filename
  Paths <- c(Paths, path) # add path to the 'Paths' vector
  
  name <- getFilename(filePath = path) # get the filename using 'getFilename' function
  Filenames <- c(Filenames, name) # add the filename to to 'Filename' vector
  
 }

No <- 1:length(Filenames)
t <- data.frame(No, Filenames)

kable(t, caption = "Table 1: Relevant BART data files to be analyzed", align = "c")  
  
```



### Data Analysis

To efficiently analyse data from all the BART countdata files, a for loop was used to run all 5 functions for all the files as shown below. The summary statistics outputs were all stored in a dataframe 'summary DF'. 

See below for code chunk:


```{r eval = TRUE, echo = TRUE}

# Create empty dataframe to store summary statistics
summaryDF <- data.frame(Filename = character(length(Paths)), 
                        Year = numeric(length(Paths)), 
                        Abundance = numeric(length(Paths)), 
                        Species_Richness = numeric(length(Paths)))

# Initiate empty vectors to store the filename, year, abundance and species richness for each file
filename <- c()
year <- c()
I_abundance <- c()
richness <- c()

# Loop through each directory and pull out the filename, year, abundance and species richness.
for (i in seq_along(Paths)){
  filename <- c(filename, getFilename(Paths[i])) 
  year <- c(year, extract_year(Paths[i]))
  data <- clean_data(Paths[i])
  
  I_abundance <- c(I_abundance, abundance(df = data))
  richness <- c(richness, species_richness(df = data))
  }

# Write the summary statistics into the summary statistics dataframe 'summaryDF'
summaryDF$Filename <- filename
summaryDF$Year <- year
summaryDF$Abundance <- I_abundance
summaryDF$Species_Richness <- richness

summaryDF %>%
  kable(caption = "Table 2 - Summary statistics for BART data",
        col.names= c("File", "Year", "Abundance", "Species Richness"), 
        align = "lccc")

```